Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-06 paper/paper/SPIE/article.tex
[WIP]

\textbf{NOTE}: Forget academia for a moment, and write like its a blog. Then go
back and add the citations and adjust the tone.

Random Matrix Theory (RMT) has obvious potential. It has helped to understand
the generalization behaviour of neural networks [add citation], can be used to
guide any principal components analysis reduction [cite: Google search: "random
matrix" "principal components"], has been used to help understand phenomena as
diverse as [include the usual citations to buses, traffic, physics, etc], and
can provide insights wherever there is a system of components interacting with
some independence [cite Cambridge RMT handbook, others].

At its most rudimentary, RMT simply describes the statistically-expected
behaviour of the eigenvalues of some classes—or \textit{ensembles}—of random
matrices under certain assumptions or conditions. A random matrix is,
generally, a matrix in which the entries can be considered to be independently
and identically distributed (\textit{iid}). For example, one of the simplest
ensembles [cite handbook] is the Gaussian Unitary Ensemble (GUE), wherein [desc
of ensemble properties] and which appears to be useful in describing
[real-world examples].

@@

For a system as complex as the human brain, RMT is thus tantalizing. It would
be quite something if we could extract some matrix reduction of brain activity,
and if the RMT description of this matrix either aligned nicely with RMT
theory, or otherwise was broadly predictively useful.

@@

\subsubsection{Baseline Features}

If RMT- or eigenvalue-based features fail to outperform simple reductions of
the fMRI data in predictive utility, it is difficult to justify the greater
computational and interpretational complexities of the former. These simple
statistical reductions help assess the relative value of RMT features.

Baseline reductions compute either a robust measure of location (mean, max,
min) or non-robust measure of location (median, \(95^{\text{th}}\) percentile,
\(5^{\text{th}}\) percentile), or a non-robust (standard deviation, range) or
robust (interquartile range, difference between \(95^{\text{th}}\) percentile
and \(5^{\text{th}}\) percentile) measure of scale summarizing across the voxel
dimension, yielding a feature of \(T\) dimensions. Exact details can be found
in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

\subsubsection{Baseline Features}
MAYBE MOVE THIS SECTION AFTER RMT AND EIGEN FEATURES?

fMRI global mean signal is

\subsubsection{Non-RMT Eigenfeatures}

@@

\subsubsection{RMT Features}
Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity, \(\Delta_3(L)\), is
calculated for any positive real value \(L < \max(\Lambda)\) as:
\[
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\]
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The spectral
rigidty for a value \(L\) is thus the \textit{average nonlinearity}

The level
number variance, \(\Sigma^2(L)\), is closely related to the spectral rigidity
\cite{mehtaRandomMatrices2004}, and is calculated as:

\[
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\]

where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\), and where \(c\), \(L\),
and \(\langle \cdot \rangle_c\) are as above \cite{guhrRandommatrixTheoriesQuantum1998a}. Both the
spectral rigidity and level number variance are also computed and investigated in this study.

>>>>

Random Matrix Theory (RMT) has obvious potential.  In small-scale physical
systems, RMT universalities have been observed in quantum chaotic systems,
complex nuclei, atoms, molecules and disordered mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.

Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}.

At its most rudimentary, RMT simply describes the statistically-expected
behaviour of the eigenvalues of some classes—or \textit{ensembles}—of random
matrices under certain assumptions or conditions. A random matrix is,
generally, a matrix in which the entries are drawn from independently
and identically distributed (\textit{iid}) distributions. For example, one of the simplest
ensembles [cite handbook] is the Gaussian Unitary Ensemble (GUE), wherein [desc
of ensemble properties] and which appears to be useful in describing
[real-world examples].

@@

For a system as complex as the human brain, RMT is thus tantalizing. If some
matrix representation of brain activity can be described, even partially, by
aspects of RMT theory, this would mean  description of this matrix either aligned nicely with
RMT theory, or otherwise was broadly predictively useful.

@@

\subsubsection{RMT Features}
\label{sec:rmt-features}

To compare the observed eigenvalue distribution to some RMT theoretical
predictions, the eigenvalues must be unfolded
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The
details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process
can be viewed as a smoothing and rescaling procedure where the originally
observed spectrum \(\lambda_1 \le \dots \le \lambda_n\) is smoothly mapped to
an unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n \approx 1\),
if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing
function is not known, and so must be approximated
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. Typically,
a polynomial is chosen \cite{abul-magdUnfoldingSpectrumChaotic2014}.

Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity,
\(\Delta_3(L)\), is calculated for any positive real value \(L <
\max(\Lambda)\) as:
\[
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\]
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or
equal to \(\lambda\), \(\langle \cdot \rangle_c\) denotes the average with
respect to all starting points \(c\), and where \(A\) and \(B\) denote the
slope and intercept, respectively, of the least squares fit of a straight line
to \(\eta(\lambda)\) on \([c, c+L]\)
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as a timeseries, the spectral rigidty for a value \(L\) can be is thus the
\textit{average nonlinearity} of all intervals of length \(L\) over the series.

The level number variance, \(\Sigma^2(L)\), or level variance, for short, is
closely related to the spectral rigidity \cite{mehtaRandomMatrices2004}, and is
calculated as:
\[
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\]
where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\),
and where \(c\), \(L\), and \(\langle \cdot \rangle_c\) are as above
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as an \textit{irregular} timeseries, the level number varaince is simply the
variance of the number of samples of all intervals of length \(L\) over the
series.

Both the spectral rigidity and level number variance are also computed and
investigated in this study. Because the computation of these statistics is
non-trivial, we developed and release the separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances.

\subsubsection{Baseline Features}
\label{sec:baselines}

MAYBE MOVE THIS SECTION AFTER RMT AND EIGEN FEATURES?

If RMT- or eigenvalue-based features fail to outperform features that are simple
summary statistics of the fMRI data in predictive utility, it is difficult to
justify the greater computational and interpretational complexities of the
former. We compute ten simple statistical reductions of the voxel time series
("timeseries features") to help assess the relative value of RMT features.

Baseline reductions use as statistical summaries: robust measures of location
(mean, max, min); non-robust measures of location (median, \(95^{\text{th}}\)
percentile, \(5^{\text{th}}\) percentile); non-robust measures of scale
(standard deviation, range); and robust measures of scale (interquartile range,
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile). These statistics are used to summarizing / reduce the fMRI data
along the voxel dimension, yielding a feature of \(T\) dimensions. E.g. the "mean"
feature is just the global mean signal. Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

\subsubsection{Non-RMT Eigenfeatures}
\label{sec:eigs-only}

@@
